Eco-evolutionary dynamics of finite populations from first principles

### 2023-04-18 ind_files/99_Appendix_D_examples.tex
Minor rephrasing

Let the solution to the equations \eqref{nD_example_det_limit} be given by $\mathbf{a}(t) = [a_1(t),\ldots,a_m(t)]^{\mathrm{T}}$. For the weak noise approximation, we can Taylor expand $A^{\pm}_i$ and use \eqref{nD_WNA_directional_derivative_for_replicator_eqns} to compute the directional derivative as:
\begin{align}
D_i &= y_iw_i(\mathbf{a}) + a_i\sum\limits_{k=1}^{m}y_k\left(\frac{\partial w_i}{\partial x_k}\bigg{|}_{\mathbf{x}=\mathbf{a}(t)}\right)\\
&= y_iw_i(\mathbf{a}) + a_i\sum\limits_{k=1}^{m}y_k\left(\frac{\partial}{\partial x_k}(r+\sum\limits_{j=1}^{m}M_{ij}x_j)\bigg{|}_{\mathbf{x}=\mathbf{a}(t)}\right)\\
&= y_iw_i(\mathbf{a}) + a_i\sum\limits_{k=1}^{m}y_kM_{ik}\\
\Rightarrow D_i &= y_iw_i(\mathbf{a}) + a_iw_i(\mathbf{y}) - ra_i\label{nD_example_directional_derivative}
\end{align}
where we have used the fact that $w_i(\mathbf{y}) = r + \sum\limits_{k=1}^{m}y_kM_{ik}$ (from \eqref{nD_example_fitness}) in the last step. Thus, in the weak noise approximation of our process, the dynamics are given by
\begin{equation}
\mathbf{x}(t) = \mathbf{a}(t) + \frac{1}{\sqrt{K}}\mathbf{y}(t)
\end{equation}
where the stochastic fluctuations $\mathbf{y}(t)$ satisfy the linear Fokker-Planck equation
\begin{equation}
\resizebox{1.1\textwidth}{!}{$\displaystyle\frac{\partial P}{\partial t}(\mathbf{y},t) = \sum\limits_{i=1}^{m}\left(-\frac{\partial}{\partial y_i}\left\{\left(y_iw_i(\mathbf{a}) + a_iw_i(\mathbf{y}) - ra_i\right)P(\mathbf{y},t)\right\}+\frac{1}{2}\left(a_i\left(\nu + \sum\limits_{j=1}^{m}T_{ij}a_j\right)\right)\frac{\partial^2}{\partial{y_i}^2}P(\mathbf{y},t)\right)$}
\end{equation}
Using \eqref{nD_example_directional_derivative} in \eqref{nD_moment_eqn_mean}, we see that the fluctuations are expected to evolve as:
\begin{equation}
\label{nD_example_moment_eqn_mean}
\frac{d}{dt}\mathbb{E}[y_i] = w_i(\mathbf{a})\mathbb{E}[y_i] + a_i\sum\limits_{k=1}^{m}M_{ik}\mathbb{E}[y_k]
\end{equation}
or, in matrix form:
\begin{equation}
\resizebox{0.93\textwidth}{!}{$\displaystyle
	\frac{d}{dt}\begin{bmatrix}
	\mathbb{E}[y_1]\\
	\mathbb{E}[y_2]\\
	\vdots\\
	\mathbb{E}[y_i]\\
	\vdots\\
	\mathbb{E}[y_m]
	\end{bmatrix}
	=
	\begin{bmatrix}
	(r + \sum\limits_{j=1}^{m}M_{1j}a_j + a_1M_{11}) & a_1M_{12} & a_1M_{13} & \dots & \dots & \dots & a_1M_{1m}\\
	a_2M_{21} & (r + \sum\limits_{j=1}^{m}M_{2j}a_j + a_2M_{22}) & a_2M_{23} & \dots & \dots & \dots & a_2M_{2m}\\
	\vdots &  & \ddots & &  & & \vdots\\
	a_{i}M_{i1} & a_iM_{i2} & a_iM_{i3} & \dots & (r + \sum\limits_{j=1}^{m}M_{ij}a_j + a_iM_{ii}) & \dots & a_iM_{im}\\
	\vdots &  &  & & & \ddots & \vdots\\
	a_mM_{m1} & a_mM_{m2} & a_mM_{m3} & \dots & \dots & \dots & (r + \sum\limits_{j=1}^{m}M_{mj}a_j + a_mM_{mm})
	\end{bmatrix}
	\begin{bmatrix}
	\mathbb{E}[y_1]\\
	\mathbb{E}[y_2]\\
	\vdots\\
	\mathbb{E}[y_i]\\
	\vdots\\
	\mathbb{E}[y_m]
	\end{bmatrix}
	$}
\end{equation}
The eigenvalues of the first matrix on the RHS will tell us whether the fixed point $\mathbb{E}[y_i] = 0 \ \forall \ i$ (the only fixed point of this system) is stable, or whether fluctuations are expected to grow (up to the point where the fluctuations are so large that the WNA is no longer valid). In the $m=2$ case,~\cite{tao_stochastic_2007} have shown that $\mathbb{E}[y_i] = 0 \ \forall \ i$ is a stable fixed point for this system iff the point $\mathbf{y}$ is an ESS (in the usual game-theoretic sense) for the matrix game defined by the payoff matrix $\mathbf{M}$.

>>>>

In frequency space, the complete mesoscopic description for the stochastic dynamics can be similarly calculated in terms of the matrices $\mathbf{M}$ and $\mathbf{T}$, and will yield:
\begin{equation}
dp_i =  N_K(t)\left( \left[(\mathbf{Mp})_i - \mathbf{p}\cdot\mathbf{Mp}\right]p_i - \frac{1}{KN_K}\left[(\mathbf{Tp})_i - \mathbf{p}\cdot\mathbf{Tp}\right]p_i\right)dt + \frac{1}{\sqrt{K}}\left(p_i(\mathbf{Tp})_i - p_i^2(\mathbf{Tp})_i\right)^{1/2}dW_t 
\end{equation}

We can also carry out the weak noise approximation for this system. Let the solution to the equations \eqref{nD_example_det_limit} be given by $\mathbf{a}(t) = [a_1(t),\ldots,a_m(t)]^{\mathrm{T}}$. For the weak noise approximation, we can Taylor expand $A^{\pm}_i$ and use \eqref{nD_WNA_directional_derivative_for_replicator_eqns} to compute the directional derivative as:
\begin{align}
D_i &= y_iw_i(\mathbf{a}) + a_i\sum\limits_{k=1}^{m}y_k\left(\frac{\partial w_i}{\partial x_k}\bigg{|}_{\mathbf{x}=\mathbf{a}(t)}\right)\\
&= y_iw_i(\mathbf{a}) + a_i\sum\limits_{k=1}^{m}y_k\left(\frac{\partial}{\partial x_k}(r+\sum\limits_{j=1}^{m}M_{ij}x_j)\bigg{|}_{\mathbf{x}=\mathbf{a}(t)}\right)\\
&= y_iw_i(\mathbf{a}) + a_i\sum\limits_{k=1}^{m}y_kM_{ik}\\
\Rightarrow D_i &= y_iw_i(\mathbf{a}) + a_iw_i(\mathbf{y}) - ra_i\label{nD_example_directional_derivative}
\end{align}
where we have used the fact that $w_i(\mathbf{y}) = r + \sum\limits_{k=1}^{m}y_kM_{ik}$ (from \eqref{nD_example_fitness}) in the last step. Thus, in the weak noise approximation of our process, the dynamics are given by
\begin{equation}
\mathbf{x}(t) = \mathbf{a}(t) + \frac{1}{\sqrt{K}}\mathbf{y}(t)
\end{equation}
where the stochastic fluctuations $\mathbf{y}(t)$ satisfy the linear Fokker-Planck equation
\begin{equation}
\resizebox{1.1\textwidth}{!}{$\displaystyle\frac{\partial P}{\partial t}(\mathbf{y},t) = \sum\limits_{i=1}^{m}\left(-\frac{\partial}{\partial y_i}\left\{\left(y_iw_i(\mathbf{a}) + a_iw_i(\mathbf{y}) - ra_i\right)P(\mathbf{y},t)\right\}+\frac{1}{2}\left(a_i\left(\nu + \sum\limits_{j=1}^{m}T_{ij}a_j\right)\right)\frac{\partial^2}{\partial{y_i}^2}P(\mathbf{y},t)\right)$}
\end{equation}
Using \eqref{nD_example_directional_derivative} in \eqref{nD_moment_eqn_mean}, we see that the fluctuations are expected to evolve as:
\begin{equation}
\label{nD_example_moment_eqn_mean}
\frac{d}{dt}\mathbb{E}[y_i] = w_i(\mathbf{a})\mathbb{E}[y_i] + a_i\sum\limits_{k=1}^{m}M_{ik}\mathbb{E}[y_k]
\end{equation}
or, in matrix form:
\begin{equation}
\resizebox{0.93\textwidth}{!}{$\displaystyle
	\frac{d}{dt}\begin{bmatrix}
	\mathbb{E}[y_1]\\
	\mathbb{E}[y_2]\\
	\vdots\\
	\mathbb{E}[y_i]\\
	\vdots\\
	\mathbb{E}[y_m]
	\end{bmatrix}
	=
	\begin{bmatrix}
	(r + \sum\limits_{j=1}^{m}M_{1j}a_j + a_1M_{11}) & a_1M_{12} & a_1M_{13} & \dots & \dots & \dots & a_1M_{1m}\\
	a_2M_{21} & (r + \sum\limits_{j=1}^{m}M_{2j}a_j + a_2M_{22}) & a_2M_{23} & \dots & \dots & \dots & a_2M_{2m}\\
	\vdots &  & \ddots & &  & & \vdots\\
	a_{i}M_{i1} & a_iM_{i2} & a_iM_{i3} & \dots & (r + \sum\limits_{j=1}^{m}M_{ij}a_j + a_iM_{ii}) & \dots & a_iM_{im}\\
	\vdots &  &  & & & \ddots & \vdots\\
	a_mM_{m1} & a_mM_{m2} & a_mM_{m3} & \dots & \dots & \dots & (r + \sum\limits_{j=1}^{m}M_{mj}a_j + a_mM_{mm})
	\end{bmatrix}
	\begin{bmatrix}
	\mathbb{E}[y_1]\\
	\mathbb{E}[y_2]\\
	\vdots\\
	\mathbb{E}[y_i]\\
	\vdots\\
	\mathbb{E}[y_m]
	\end{bmatrix}
	$}
\end{equation}
The eigenvalues of the first matrix on the RHS will tell us whether the fixed point $\mathbb{E}[y_i] = 0 \ \forall \ i$ (the only fixed point of this system) is stable, or whether fluctuations are expected to grow (up to the point where the fluctuations are so large that the WNA is no longer valid). In the $m=2$ case,~\cite{tao_stochastic_2007} have shown that $\mathbb{E}[y_i] = 0 \ \forall \ i$ is a stable fixed point for this system iff the point $\mathbf{y}$ is an ESS (in the usual game-theoretic sense) for the matrix game defined by the payoff matrix $\mathbf{M}$.

### 2023-04-18 ind_files/99_Appendix_E_lemma.tex
Minor rephrasing

\begin{lemma}
Let $m \in \mathbb{N}$. Let $W^{(1)}_t, W^{(2)}_t, \ldots, W^{(m)}_t$ be $m$ independent one-dimensional Wiener processes. Let $g_1(x), g_2(x), \ldots, g_m(x)$ be $m$ `nice' ($L^2(\mathbb{R})$, Lipschitz, \emph{etc}.) real functions. Let
\begin{equation*}
	dX_t = \sum\limits_{i=1}^{m} g_i(X_t)dW^{(i)}_t
\end{equation*}
Then, we can always find a \emph{single} one-dimensional Wiener process $W_t$ (on the same probability space) such that
\begin{equation*}
	dX_t = \left(\sum\limits_{i=1}^{m} g^2_i(X_t)\right)^{1/2}dW_t
\end{equation*}
\end{lemma}
\begin{proof}
It suffices to prove the $m=2$ case.\\
Let $dX_t$ = $g_1(X_t) dW^{(1)}_t + g_2(X_t)dW^{(2)}_t$. Let us consider the \emph{two}-dimensional process $\mathbf{W}_t = [W^{(1)}_t, W^{(2)}_t]^{\mathrm{T}}$ on $\mathbb{R}^2$. Let us define a function $G:\mathbb{R} \to \mathbb{R}^2$ as
\begin{equation}
\label{lemma_G_defn}
G(x) = \frac{1}{\sqrt{g_1^2(x) + g_2^2(x)}}\begin{bmatrix}g_1(x) \\ g_2(x)\end{bmatrix}
\end{equation}
Now, by definition, we have
\begin{equation}
	\label{lemma_G_integral}
\int\limits_{0}^{t}G(X_s) \cdot d\mathbf{W}_s = \int\limits_{0}^{t}\frac{g_1(X_s)}{\sqrt{g_1^2(X_s) + g_2^2(X_s)}} dW^{(1)}_t + \int\limits_{0}^{t}\frac{g_2(X_s)}{\sqrt{g_1^2(X_s) + g_2^2(X_s)}}dW^{(2)}_s
\end{equation}
By a simple corollary of the It\^o isometry, we can calculate the quadratic variation of $\int G \cdot d\mathbf{W}$ as
\begin{align}
\bigg\langle \int\limits G(X_s) \cdot d\mathbf{W}_s \bigg\rangle_t &= \int\limits_{0}^{t} \bignorm{ G(X_s) }^2 d \langle \mathbf{W} \rangle_s \nonumber\\
&= \int\limits_{0}^{t}  \frac{1}{g_1^2 + g_2^2} \cdot (g_1^2 + g_2^2) ds = \int\limits_{0}^{t} ds  = t \label{lemma_ito_isometry_step}
\end{align}
Further, since $\int G \cdot d\mathbf{W}$ is a stochastic integral, it is guaranteed to be a continuous square-integrable martingale.  But, by L\'evy's characterization of Brownian motion, the only continuous martingale $M_t$ that satisfies $\langle M \rangle_t = t$ is the Brownian motion. Thus, from equation \eqref{lemma_ito_isometry_step}, we are led to conclude that there is a one-dimensional Wiener process $W_t$ on the same probability space such that we can write
\begin{equation}
	\label{lemma_G_is_W}
	G(X_t) \cdot d\mathbf{W}_t = dW_t
\end{equation}
We can now use equation \eqref{lemma_G_integral} on the LHS of equation \eqref{lemma_G_is_W} to write
\begin{align}
	\frac{g_1(X_t)}{\sqrt{g_1^2(X_t) + g_2^2(X_t)}} dW^{(1)}_t + \frac{g_2(X_t)}{\sqrt{g_1^2(X_t) + g_2^2(X_t)}}dW^{(2)}_t &= dW_t\\
	\Rightarrow g_1 dW^{(1)}_t + g_2dW^{(2)}_t &= \sqrt{g_1^2(X_t) + g_2^2(X_t)}dW_t
\end{align}
By definition of our original process $X_t$, we can now conclude that
\begin{equation}
	dX_t = \sqrt{g_1^2(X_t) + g_2^2(X_t)}dW_t
\end{equation}
thus completing the proof.
\end{proof}
Using this lemma, we can now calculate the stochastic integral terms of our equations. For equation \eqref{nD_stochastic_Price}, the stochastic analog of the Price equation, we have:
\begin{equation}
dW_{\overline{f}} = \sum\limits_{i=1}^{m}\left(f_i-\overline{f}\right)\sqrt{A_i^+}dW_{t}^{(i)} = \left(\sum\limits_{i=1}^{m}\left(f_i-\overline{f}\right)^2A_i^+\right)^{1/2}dW_t
\end{equation}
where $dW_t$ is now a single one-dimensional Wiener process. This is precisely the term calculated in equation \eqref{intermediate_2_for_variances} (barring the $1/KN_K^2$ pre-factor), and thus the stochastic term for the mean value is given by:
\begin{equation}
	\label{nD_stoch_Price_noise_term_alt_representation}
	dW_{\overline{f}} = \sqrt{N_K(t) \left(\textrm{Cov}(\tau,\left(f - \overline{f}\right)^2) + \overline{\tau}\sigma^2_{f} +\mu \sum\limits_{i=1}^{m}\left(f_i - \overline{f}\right)^2Q_i(\mathbf{p})\right)}dW_t
\end{equation}

>>>>

\begin{lemma}
Let $m \in \mathbb{N}$. Let $W^{(1)}_t, W^{(2)}_t, \ldots, W^{(m)}_t$ be $m$ independent one-dimensional Wiener processes. Let $g_1(x), g_2(x), \ldots, g_m(x)$ be $m$ `nice' ($L^2(\mathbb{R})$, Lipschitz, \emph{etc}.) real functions. Let
\begin{equation*}
	dX_t = \sum\limits_{i=1}^{m} g_i(X_t)dW^{(i)}_t
\end{equation*}
Then, we can always find a \emph{single} one-dimensional Wiener process $W_t$ (on the same probability space) such that
\begin{equation*}
	dX_t = \left(\sum\limits_{i=1}^{m} g^2_i(X_t)\right)^{1/2}dW_t
\end{equation*}
\end{lemma}
\begin{proof}
It suffices to prove the $m=2$ case.\\
Let $dX_t$ = $g_1(X_t) dW^{(1)}_t + g_2(X_t)dW^{(2)}_t$. Let us consider the \emph{two}-dimensional process $\mathbf{W}_t = [W^{(1)}_t, W^{(2)}_t]^{\mathrm{T}}$ on $\mathbb{R}^2$. Let us define a function $G:\mathbb{R} \to \mathbb{R}^2$ as
\begin{equation}
\label{lemma_G_defn}
G(x) = \frac{1}{\sqrt{g_1^2(x) + g_2^2(x)}}\begin{bmatrix}g_1(x) \\ g_2(x)\end{bmatrix}
\end{equation}
Now, by definition, we have
\begin{equation}
	\label{lemma_G_integral}
\int\limits_{0}^{t}G(X_s) \cdot d\mathbf{W}_s = \int\limits_{0}^{t}\frac{g_1(X_s)}{\sqrt{g_1^2(X_s) + g_2^2(X_s)}} dW^{(1)}_t + \int\limits_{0}^{t}\frac{g_2(X_s)}{\sqrt{g_1^2(X_s) + g_2^2(X_s)}}dW^{(2)}_s
\end{equation}
By a simple corollary of the It\^o isometry, we can calculate the quadratic variation of $\int G \cdot d\mathbf{W}$ as
\begin{align}
\bigg\langle \int\limits G(X_s) \cdot d\mathbf{W}_s \bigg\rangle_t &= \int\limits_{0}^{t} \bignorm{ G(X_s) }^2 d \langle \mathbf{W} \rangle_s \nonumber\\
&= \int\limits_{0}^{t}  \frac{1}{g_1^2 + g_2^2} \cdot (g_1^2 + g_2^2) ds = \int\limits_{0}^{t} ds  = t \label{lemma_ito_isometry_step}
\end{align}
Further, since $\int G \cdot d\mathbf{W}$ is a stochastic integral, the process $M_t = \int_0^t G(X_s) \cdot d\mathbf{W}_s$ is guaranteed to be a continuous square-integrable martingale.  But, by L\'evy's characterization of Brownian motion, the only continuous martingale $M_t$ that satisfies $\langle M \rangle_t = t$ is the Brownian motion. Thus, from equation \eqref{lemma_ito_isometry_step}, we are led to conclude that there is a one-dimensional Wiener process $W_t$ on the same probability space such that we can write
\begin{equation}
	\label{lemma_G_is_W}
	G(X_t) \cdot d\mathbf{W}_t = dW_t
\end{equation}
We can now use equation \eqref{lemma_G_integral} on the LHS of equation \eqref{lemma_G_is_W} to write
\begin{align}
	\frac{g_1(X_t)}{\sqrt{g_1^2(X_t) + g_2^2(X_t)}} dW^{(1)}_t + \frac{g_2(X_t)}{\sqrt{g_1^2(X_t) + g_2^2(X_t)}}dW^{(2)}_t &= dW_t\\
	\Rightarrow g_1 dW^{(1)}_t + g_2dW^{(2)}_t &= \sqrt{g_1^2(X_t) + g_2^2(X_t)}dW_t
\end{align}
By definition of our original process $X_t$, we can now conclude that
\begin{equation}
	dX_t = \sqrt{g_1^2(X_t) + g_2^2(X_t)}dW_t
\end{equation}
thus completing the proof.
\end{proof}
Using this lemma, we can now calculate the stochastic integral terms of our equations. For equation \eqref{nD_stochastic_Price}, the stochastic analog of the Price equation, we have:
\begin{equation}
dW_{\overline{f}} = \sum\limits_{i=1}^{m}\left(f_i-\overline{f}\right)\sqrt{A_i^+}dW_{t}^{(i)} = \left(\sum\limits_{i=1}^{m}\left(f_i-\overline{f}\right)^2A_i^+\right)^{1/2}dW_t
\end{equation}
where $dW_t$ is now a single one-dimensional Wiener process. This is precisely the term calculated in equation \eqref{intermediate_2_for_variances} (barring the $1/KN_K^2$ pre-factor), and thus the stochastic term for the mean value is given by:
\begin{equation}
	\label{nD_stoch_Price_noise_term_alt_representation}
	dW_{\overline{f}} = \sqrt{N_K(t) \left(\textrm{Cov}(\tau,\left(f - \overline{f}\right)^2) + \overline{\tau}\sigma^2_{f} +\mu \sum\limits_{i=1}^{m}\left(f_i - \overline{f}\right)^2Q_i(\mathbf{p})\right)}dW_t
\end{equation}
